---
title: "Statistical Models"
subtitle: "Appendix A"
from: markdown+emoji
author: 
    - name: Dr. Silvio Fanzon
      id: sf
      email: S.Fanzon@hull.ac.uk
      url: https://www.silviofanzon.com
      affiliations: University of Hull
---



::: {.content-hidden}
$
{{< include macros.tex >}}
$
:::




# Appendix A: <br>Probability Revision {background-color="#cc0164" visibility="uncounted"}

::: footer

<div color="#cc0164">  </div>

:::





## Introduction {.smaller}

- In this Appendix we review some fundamental Probability notions
- You saw these in the Y1 module **Introduction to Probability & Statistics**
- We will adopt a slightly more mature mathematical approach
- Remember: The mathematical description might look (a bit) different, but the concepts are the same

**Topics reviewed**:
    
::: {.column width="38%"}
    
- Sample space
- Events
- Probability measure
- Conditional probability
- Events independence

:::

::: {.column width="58%"}

- Random Variable (Discrete and Continuous)
- Distribution
- cdf, pmf, pdf
- Expected value and Variance

:::




## Outline of Appendix A

1. Probability space
2. Random variables
3. Expected value




# Part 1: <br>Probability space {background-color="#cc0164" visibility="uncounted"}

::: footer

<div color="#cc0164">  </div>

:::





## Sample space {.smaller}


::: Definition
### Sample space

A set $\Omega$ of all possible outcomes of some experiment

:::


**Examples**:

- Coin toss: results in Heads $= H$ and Tails $= T$ 
$$
\Omega = \{ H, T \}
$$

- Student grade for *Statistical Models*: a number between $0$ and $100$
$$
\Omega = \{ x \in \mathbb{R} \, \colon \, 0 \leq x \leq 100 \} = [0,100]
$$






## Events {.smaller}


::: Definition
### Event

A subset $E$ of the sample space $\Omega$ (including $\emptyset$ and $\Omega$ itself)

:::


**Operations with events**:

- **Union** of two events $A$ and $B$
$$
A \cup B := \{ x \in \Omega \colon x \in A \, \text{ or } \, x \in B \}
$$

- **Intersection** of two events $A$ and $B$
$$
A \cap B := \{ x \in \Omega \colon x \in A \, \text{ and } \, x \in B \}
$$




## Events {.smaller}


**More Operations with events**:

- **Complement** of an event $A$
$$
A^c  := \{ x \in \Omega \colon x \notin A  \}
$$


- **Infinite Union** of a family of events $A_i$ with $i \in I$

$$
\bigcup_{i \in I} A_i := \{ x \in \Omega \colon x \in A_i \, \text{ for some } \, i \in I \}
$$

- **Infinite Intersection** of a family of events $A_i$ with $i \in I$

$$
\bigcap_{i \in I} A_i := \{ x \in \Omega \colon x \in A_i \, \text{ for all } \, i \in I \}
$$





## Events {.smaller}

**Example**: Consider sample space and events
$$
\Omega := (0,1] \,, \quad 
A_i = \left[\frac{1}{i} , 1 \right] \,, \quad i \in \mathbb{N}
$$
Then
$$
\bigcup_{i \in I} A_i = (0,1] \,, \quad 
\bigcap_{i \in I} A_i = \{ 1 \}
$$





## Events {.smaller}

::: Definition
### Disjoint

Two events $A$ and $B$ are **disjoint** if
$$
A \cap B = \emptyset
$$
Events $A_1, A_2, \ldots$ are **pairwise disjoint** if
$$
A_i \cap A_j = \emptyset \,, \quad \forall \, i \neq j
$$
:::





## Events {.smaller}

::: Definition
### Partition

The collection of events  $A_1, A_2, \ldots$ is a **partition** of $\Omega$ if

1. $A_1, A_2, \ldots$ are pairwise disjoint
2. $\Omega = \cup_{i=1}^\infty A_i$

:::






## What's a Probability? {.smaller}


- To each event $E \subset \Omega$ we would like to associate a number
$$
P(E) \in [0,1]
$$

- The number $P(E)$ is called the **probability** of $E$

- The number $P(E)$ models the **frequency of occurrence** of $E$:

    * $P(E)$ small means $E$ has low chance of occurring
    * $P(E)$ large means $E$ has high chance of occurring

- **Technical issue**:

    * One cannot associate a number $P(E)$ for all events in $\Omega$
    * Probability function $P$ only defined for a **smaller** family of events
    * Such family of events is called $\sigma$-algebra






## $\sigma$-algebras {.smaller}


::: Definition
### sigma-algebra

Let $\mathcal{B}$ be a collection of events. We say that $\mathcal{B}$ is a $\sigma$-algebra if

1. $\emptyset \in \mathcal{B}$
2. If $A \in \mathcal{B}$ then $A^c \in \mathcal{B}$
3. If $A_1,A_2 , \ldots \in \mathcal{B}$ then $\cup_{i=1}^\infty A_i \in \mathcal{B}$


:::



**Remarks**:

- Since $\emptyset \in \mathcal{B}$ and $\emptyset^c = \Omega$, we deduce that
$\Omega \in \mathcal{B}$

- Thanks to DeMorgan's Law we have that
$$
A_1,A_2 , \ldots \in \mathcal{B} \quad \implies \quad \cap_{i=1}^\infty A_i \in \mathcal{B}
$$






## $\sigma$-algebras {.smaller}
### Examples

Suppose $\Omega$ is any set: 

- Then
$$
\mathcal{B} = \{ \emptyset, \Omega \}
$$
is a $\sigma$-algebra


- The power set of $\Omega$
$$
\mathcal{B} = \operatorname{Power} (\Omega) := \{ A \colon A \subset \Omega \}
$$
is a $\sigma$-algebra






## $\sigma$-algebras {.smaller}
### Examples


- If $\Omega$ has $n$ elements then $\mathcal{B} = \operatorname{Power} (\Omega)$ contains $2^n$ sets

- If $\Omega = \{ 1,2,3\}$ then
\begin{align*}
\mathcal{B} = \operatorname{Power} (\Omega) = \big\{ & \{1\} , \, \{2\}, \, \{3\}  \\
                                       &  \{1,2\} , \, \{2,3\}, \, \{1,3\}      \\
                                       & \emptyset , \{1,2,3\}   \big\}
\end{align*}

- If $\Omega$ is **uncountable** then the power set of $\Omega$ is not easy to describe






## Lebesgue $\sigma$-algebra {.smaller}


::: Question

$\mathbb{R}$ is uncountable. Which $\sigma$-algebra do we consider?

:::


::: Definition 
### Lebesgue sigma-algebra


The Lebesgue $\sigma$-algebra on $\mathbb{R}$ is the smallest $\sigma$-algebra $\mathcal{L}$ containing all sets of the form
$$
(a,b) \,, \quad (a,b] \,, \quad [a,b) \,, \quad [a,b] 
$$
for all $a,b \in \mathbb{R}$

:::



## Lebesgue $\sigma$-algebra {.smaller}


::: Important

Therefore the events of $\R$ are 

- Intervals
- Unions and intersection of intervals
- Countable Unions and intersection of intervals

:::


::: Warning 

- I only told you that the Lebsesgue $\sigma$-algebra $\mathcal{L}$ **exists**
- Explicitly showing that $\mathcal{L}$ exists is not easy, see [@rosenthal]

:::




## Probability measure {.smaller}

Suppose given:

- $\Omega$ sample space 
- $\mathcal{B}$ a $\sigma$-algebra on $\Omega$

::: Definition
### Probability measure

A **probability measure** on $\Omega$ is a map
$$
P \colon \mathcal{B} \to [0,1]
$$
such that the **Axioms of Probability** hold

1. $P(\Omega) = 1$
2. If $A_1, A_2,\ldots$ are pairwise disjoint then
$P\left( \bigcup_{i=1}^\infty  A_i \right)  = \sum_{i=1}^\infty P(A_i)$

:::




## Properties of Probability {.smaller}

Let $A, B \in \mathcal{B}$. As a consequence of the Axioms of Probability:


1. $P(\emptyset) = 0$
2. If $A$ and $B$ are disjoint then
$$
P(A \cup B) = P(A) + P(B)
$$
3. $P(A^c) = 1 - P(A)$
4. $P(A) = P(A \cap B) + P(A \cap B^c)$
5. $P(A\cup B) = P(A) + P(B) - P(A \cap B)$
6. If $A \subset B$ then
$$
P(A) \leq P(B)
$$





## Properties of Probability {.smaller} 

7. Suppose $A$ is an event and $B_1,B_2, \ldots$ a partition of $\Omega$. Then
$$
P(A) = \sum_{i=1}^\infty  P(A \cap B_i)
$$
8. Suppose $A_1,A_2, \ldots$ are events. Then
$$
P\left( \bigcup_{i=1}^\infty  A_i \right)  \leq  \sum_{i=1}^\infty P(A_i)
$$




## Example: Fair Coin Toss {.smaller}

- The sample space for **coin toss** is $\Omega = \{ H, T \}$

- We take as $\sigma$-algebra the power set of $\Omega$
$$
\mathcal{B} = \{  \emptyset , \, \{H\}  , \, \{T\} , \, \{H,T\}   \}
$$

- We suppose that the coin is fair

    * This means $P \colon \mathcal{B} \to [0,1]$ satisfies
    $$
    P(\{H\}) = P(\{T\})
    $$
    * Assuming the above we get
    $$
    1 = P(\Omega) = P(\{H\} \cup \{T\}) = 
    P(\{H\}) + P(\{T\}) = 2 P(\{H\})
    $$
    * Therefore 
    $P(\{H\}) = P(\{T\}) = \frac12$





## Conditional Probability {.smaller}

::: Definition
### Conditional Probability

Let $A,B$ be events in $\Omega$ with 
$$
P(B)>0 
$$ 
The **conditional probability** of $A$ given $B$ is
$$
P(A|B) := \frac{P(A \cap B)}{P(B)} 
$$

:::




## Conditional Probability {.smaller}
### Intuition

The conditional probability
$$
P(A|B) = \frac{P( A \cap B)}{P(B)}
$$ 
represents the probability of $A$, knowing that $B$ has happened:

- If $B$ has happened, then $B$ is the *new sample space* 
- Therefore $A \cap B^c$ cannot happen, and we are only interested in $A \cap B$
- Hence it makes sense to define $P(A|B) \propto P(A \cap B)$
- We divide $P(A\cap B)$ by $P(B)$ so that $P(A|B) \in [0,1]$ is still a probability
- The function $A \mapsto P(A|B)$ is a probability measure on $\Omega$




## Bayes' Rule {.smaller}


- For two events $A$ and $B$ is holds

$$
P(A | B ) = P(B|A) \frac{P(A)}{P(B)}
$$


- Given a partition $A_1, A_2, \ldots$ of the sample space we have

$$
P(A_i | B ) = \frac{ P(B|A_i) P(A_i)}{\sum_{j=1}^\infty P(B | A_j) P(A_j)}
$$





## Independence {.smaller}


::: Definition

Two events $A$ and $B$ are **independent** if 
$$
P(A \cap B) = P(A)P(B)
$$
A collection of events $A_1 , \ldots ,A_n$ are **mutually independent** if
for any subcollection $A_{i_1}, \ldots, A_{i_k}$ it holds
$$
P \left(  \bigcap_{j=1}^k A_j  \right) = \prod_{j=1}^k  P(A_{i_j})
$$

:::





# Part 2: <br>Random variables {background-color="#cc0164" visibility="uncounted"}

::: footer

<div color="#cc0164">  </div>

:::



## Random Variables {.smaller}
### Motivation

- Consider the experiment of flipping a coin $50$ times
- The sample space consists of $2^{50}$ elements
- Elements are vectors of $50$ entries recording the outcome $H$ or $T$ of each flip
- This is a very large sample space!


Suppose we are only interested in
$$
X = \text{ number of } \, H \, \text{ in } \, 50 \, \text{flips}
$$

- Then the new sample space is the set of integers
$\{ 0,1,2,\ldots,50\}$
- This is much smaller!
- $X$ is called a Random Variable




## Random Variables {.smaller}


Assume given

- $\Omega$ sample space
- $\mathcal{B}$ a $\sigma$-algebra of events on $\Omega$
- $P \colon \mathcal{B} \to [0,1]$ a probability measure


::: Definition
### Random variable

A function $X \colon \Omega \to \mathbb{R}$

:::


We will abbreviate Random Variable with rv





## Random Variables {.smaller}
### Technical remark 


::: Definition
### Random variable

A [**measurable**]{style="color:#cc0164;"}
function $X \colon \Omega \to \mathbb{R}$

:::


**Technicality**: 
$X$ is a [**measurable**]{style="color:#cc0164;"}
function if
$$
\{ X \in I \} := \{ \omega \in \Omega \colon X(\omega) \in I \} \in \mathcal{B} \,, \quad \forall \, I \in \mathcal{L}
$$
where 

- $\mathcal{L}$ is the Lebsgue $\sigma$-algebra on $\mathbb{R}$
- $\mathcal{B}$ is the given $\sigma$-algebra on $\Omega$




## Random Variables {.smaller}
### Notation

- In particular $I \in \mathcal{L}$ can be of the form
$$
(a,b) \,, \quad (a,b] \,, \quad [a,b) \,, \quad [a,b] \,, \quad 
\forall \, a, b \in \mathbb{R}
$$

- In this case the set 
$$
\{X \in I\} \in \mathcal{B}
$$ 
is denoted by, respectively:
$$
\{ a < X < b \} \,, \quad \{ a < X \leq b \} \,, \quad \{ a \leq X < b \} \,, \quad \{ a \leq X \leq b \}
$$

- If $a=b=x$ then $I=[x,x]=\{x\}$. Then we denote
$$
\{X \in I\} = \{X = x\}
$$



## Distribution {.smaller}
### Why do we require measurability?

**Answer**: Because it allows to define a new probability measure on $\mathbb{R}$


::: Definition
### Distribution

The **distribution** of a random variable $X \colon \Omega \to \mathbb{R}$ is the probability measure on $\mathbb{R}$
$$
P_X \colon \mathcal{L} \to [0,1] \,, \quad P_X (I) := P \left( \{X \in I\} \right) \,, \,\, \forall \, I \in  \mathcal{L}
$$

:::


**Note**: 

- One can show that $P_X$ satisfies the *Probability Axioms*
- Thus $P_X$ is a probability measure on $\mathbb{R}$
- In the future we will denote 
$P \left( X \in I \right) := P \left( \{X \in I\} \right)$







## Distribution {.smaller}
### Why is the distribution useful?

**Answer**: Because it allows to define a random variable $X$

- by specifying the distribution values
$$
P \left( X \in I \right) 
$$
- rather than defining an explicit function $X \colon \Omega \to \mathbb{R}$


**Important**: More often than not

- We care about the distribution of $X$
- We do not care about how $X$ is defined





## Example - Three coin tosses {.smaller}


- Sample space $\Omega$ given by the below values of $\omega$

+:---------:+:----:+:---:+:---:+:---:+:---:+:---:+:---:+:---:+
| $\omega$  | HHH  | HHT | HTH | THH | TTH | THT | HTT | TTT |
+-----------+------+-----+-----+-----+-----+-----+-----+-----+


- The probability of each outcome is the same
$$
P(\omega) = \frac{1}{2} \times \frac{1}{2} \times \frac{1}{2} = \frac{1}{8} \,, \quad \forall \, \omega \in \Omega
$$


- Define the random variable $X \colon \Omega \to \R$ by
$$
X(\omega) :=  \text{ Number of H in }  \omega
$$

|           |      |     |     |     |     |     |     |     |
|:---------:|:----:|:---:|:---:|:---:|:---:|:---:|:---:|:---:|
| $\omega$  | HHH  | HHT | HTH | THH | TTH | THT | HTT | TTT |
|$X(\omega)$|  $3$ | $2$ | $2$ | $2$ | $1$ | $1$ | $1$ | $0$ | 




## Example - Three coin tosses {.smaller}


- Recall the definition of $X$

|           |      |     |     |     |     |     |     |     |
|:---------:|:----:|:---:|:---:|:---:|:---:|:---:|:---:|:---:|
| $\omega$  | HHH  | HHT | HTH | THH | TTH | THT | HTT | TTT |
|$X(\omega)$|  $3$ | $2$ | $2$ | $2$ | $1$ | $1$ | $1$ | $0$ | 


- The range of $X$ is $\{0,1,2,3\}$

- Hence the only interesting values of $P_X$ are
$$
P(X=0) \,, \quad 
P(X=1) \,, \quad 
P(X=2) \,, \quad 
P(X=3)
$$




## Example - Three coin tosses {.smaller}


- Recall the definition of $X$

|           |      |     |     |     |     |     |     |     |
|:---------:|:----:|:---:|:---:|:---:|:---:|:---:|:---:|:---:|
| $\omega$  | HHH  | HHT | HTH | THH | TTH | THT | HTT | TTT |
|$X(\omega)$|  $3$ | $2$ | $2$ | $2$ | $1$ | $1$ | $1$ | $0$ | 


- We compute
\begin{align*}
P(X=0) & = P(TTT) = \frac{1}{8} \\
P(X=1) & = P(TTH) + P(THT) + P(HTT) = \frac{3}{8} \\
P(X=2) & = P(HHT) + P(HTH) + P(THH) = \frac{3}{8} \\
P(X=3) & = P(HHH) = \frac{1}{8}
\end{align*}




## Example - Three coin tosses {.smaller}


- Recall the definition of $X$

|           |      |     |     |     |     |     |     |     |
|:---------:|:----:|:---:|:---:|:---:|:---:|:---:|:---:|:---:|
| $\omega$  | HHH  | HHT | HTH | THH | TTH | THT | HTT | TTT |
|$X(\omega)$|  $3$ | $2$ | $2$ | $2$ | $1$ | $1$ | $1$ | $0$ | 


- The distribution of $X$ is summarized in the table below

|           |               |               |              |               |
|:---------:|:-------------:|:-------------:|:------------:|:-------------:|
| $x$       | $0$           | $1$           | $2$          |     $3$       |
|$P(X=x)$   | $\frac{1}{8}$ | $\frac{3}{8}$ | $\frac{3}{8}$| $\frac{1}{8}$ | 
: {tbl-colwidths="[40,15,15,15,15]"}






## Cumulative Distribution Function {.smaller}


**Recall**: The **distribution** of a rv $X \colon \Omega \to \mathbb{R}$ is the probability measure on $\mathbb{R}$
$$
P_X \colon \mathcal{L} \to [0,1] \,, \quad P_X (I) := P \left( X \in I \right) \,, \,\, \forall \, I \in  \mathcal{L}
$$

::: Definition
### cdf

The **cumulative distribution function** or **cdf** of a rv $X \colon \Omega \to \mathbb{R}$ is
$$
F_X \colon \mathbb{R} \to \mathbb{R} \,, \quad 
F_X(x) := P_X (X \leq x) 
$$

:::


## Cumulative Distribution Function {.smaller}
### Intuition

- $F_X$ is the *primitive* of $P_X$:
    * Recall from Analysis: The primitive of a continuous function $g \colon \R \to \R$ is 
    $$
    G(x):=\int_{-\infty}^x g(y) \,dy
    $$
    * Note that $P_X$ is not a function but a **distribution**
    * However the definition of cdf as a primitive still makes sense


- $P_X$ will be the *derivative* of $F_X$ - In a suitable generalized sense
    * Recall from Analysis: Fundamental Theorem of Calculus says 
    $G'(x)=g(x)$
    * Since $F_X$ is the primitive of $P_X$, it will still hold
    $F_X'=P_X$ in the **sense of distributions**
    




## Distribution Function {.smaller}
### Example


- Consider again 3 coin tosses and the rv
$X(\omega) :=  \text{ Number of H in }  \omega$

- We computed that the distribution $P_X$ of $X$ is

|           |               |               |              |               |
|:---------:|:-------------:|:-------------:|:------------:|:-------------:|
| $x$       | $0$           | $1$           | $2$          |     $3$       |
|$P(X=x)$   | $\frac{1}{8}$ | $\frac{3}{8}$ | $\frac{3}{8}$| $\frac{1}{8}$ | 
: {tbl-colwidths="[40,15,15,15,15]"}


::: {.column width="38%"}

- One can compute
$$
F_X(x) = 
\begin{cases}
0  & \text{if } x < 0 \\
\frac{1}{8}  & \text{if } 0 \leq x < 1 \\
\frac{1}{2}  & \text{if } 1 \leq x < 2 \\
\frac{7}{8}  & \text{if } 2 \leq x < 3 \\
1            & \text{if } 3 \leq x
\end{cases}
$$

:::


::: {.column width="56%"}

- For example
\begin{align*}
F_X(2.1) & = P(X \leq 2.1) \\
         & = P(X=0,1 \text{ or } 2) \\
         & = P(X=0) + P(X=1) + P(X=2) \\
         & = \frac{1}{8} + \frac{3}{8} + \frac{3}{8} = \frac{7}{8}
\end{align*}

:::





## Cumulative Distribution Function {.smaller}
### Example


```{r}
# Define the step function
step_function <- function(x) {
  if (x < 0) {
    return(0)
  } else if (x < 1) {
    return(1/8)
  } else if (x < 2) {
    return(1/2)
  } else if (x < 3) {
    return(7/8)
  } else {
    return(1)
  }
}

# Generate x values
x_values <- seq(-1, 4, length.out = 1000)

# Calculate corresponding y values using the step function
y_values <- sapply(x_values, step_function)

# Create a data frame for ggplot
plot(x_values, 
    y_values, 
    type = 's',
    xlab = "",
    ylab = "")

mtext("x", side=1, line=3, cex=2)
mtext("F_X(x)", side=2, line=2.5, cex=2)

```



::: {.column width="43%"}

- Plot of $F_X$: it is a **step function**
- $F_X'=0$ except at $x=0,1,2,3$
- $F_X$ jumps at $x=0,1,2,3$ 

:::

::: {.column width="56%"}

- Size of jump at $x$ is $P(X=x)$ 
- $F_X'=P_X$ in the sense of **distributions**
<br> (Advanced analysis concept - not covered)

:::




## Discrete Random Variables {.smaller}

In the previous example:

- The cdf $F_X$ had **jumps**
- Hence $F_X$ was **discountinuous**
- We take this as definition of **discrete** rv



::: Definition

$X \colon \Omega \to \mathbb{R}$ is **discrete** if $F_X$ has jumps

:::





## Probability Mass Function {.smaller}

- In this slide $X$ is a **discrete** rv
- Therefore $F_X$ has **jumps**

::: Definition

The **Probability Mass Function** or **pmf** of a discrete rv $X$ is
$$
f_X \colon \mathbb{R} \to \mathbb{R} \,, \quad 
f_X(x) := P(X = x) 
$$

:::



## Probability Mass Function {.smaller}

::: Proposition

The **pmf** $f_X(x) = P(X=x)$ can be used to

- compute probabilities
    $$
    P(a \leq X \leq b) = \sum_{k = a}^b f_X (k) \,, \quad \forall \, a,b \in 
    \mathbb{Z} \,, \,\, a \leq b 
    $$
- compute the **cdf**
    $$
    F_X(x) = P(X \leq x) = \sum_{k=-\infty}^x f_X(k)
    $$

:::





## Example 1 - Discrete RV {.smaller}

- Consider again 3 coin tosses and the RV
$X(\omega) :=  \text{ Number of H in }  \omega$

- The pmf of $X$ is $f_X(x):=P(X=x)$, which we have already computed

|           |               |               |              |               |
|:---------:|:-------------:|:-------------:|:------------:|:-------------:|
| $x$       | $0$           | $1$           | $2$          |     $3$       |
|$f_X(x)= P(X=x)$ | $\frac{1}{8}$ | $\frac{3}{8}$ | $\frac{3}{8}$| $\frac{1}{8}$ | 
: {tbl-colwidths="[40,15,15,15,15]"}



```{r}
# Example data
values <- c(0, 1, 2, 3)
probabilities <- c(1/8, 3/8, 3/8, 1/8)

# Plot PMF
plot(values, 
    probabilities, 
    xlab = "",
    ylab = "",
    xaxt='n',
    type="h", 
    lwd=4, 
    col="blue")

axis(1, at = values,labels = T)
mtext("x", side=1, line=3, cex=2)
mtext("f_X(x)", side=2, line=2.5, cex=2)

points(values, probabilities, col = "red", pch = 16)

grid()
```







## Example 2 - Geometric Distribution {.smaller}

- Suppose $p \in (0,1)$ is a given probability of **success**
- Hence $1-p$ is probability of **failure**
- Consider the random variable 
$$
X = \text{ Number of attempts to obtain first success}
$$
- Since each trial is independent, the pmf of $X$ is
$$
f_X (x) = P(X=x) = (1-p)^{x-1} p \,, \quad \forall \, x \in \mathbb{N}
$$
- This is called **geometric distribution**



## Example 2 - Geometric Distribution {.smaller}

- We want to compute the cdf of $X$: For $x \in \mathbb{N}$ with $x > 0$
\begin{align*}
F_X(x) & = P(X \leq x) = \sum_{k=1}^x P(X=k) = \sum_{k=1}^x f_X(k) \\
       & = \sum_{k=1}^x (1-p)^{k-1} p 
         = \frac{1-(1-p)^x}{1-(1-p)} p = 1 - (1-p)^x 
\end{align*}
where we used the formula for the sum of geometric series:
$$
\sum_{k=1}^x t^{k-1} = \frac{1-t^x}{1-t} \,, \quad t \neq 1 
$$



## Example 2 - Geometric Distribution {.smaller}


- $F_X$ is flat between two consecutive natural numbers:
\begin{align*}
F_X(x+k) & = P(X \leq x+k) \\
         & = P(X \leq x) \\
         & = F_X(x) 
\end{align*} 
for all $x \in \mathbb{N}, k \in [0,1)$

- Therefore $F_X$ has **jumps** and $X$ is **discrete**





## Continuous Random Variables {.smaller} 

**Recall**: $X$ is **discrete** if $F_X$ has **jumps**


::: Definition
### Continuous Random Variable

$X \colon \Omega \to \mathbb{R}$ is **continuous** if 
$F_X$ is **continuous**

:::







## Probability Mass Function? {.smaller}

- Suppose $X$ is a **continuous** rv
- Therefore $F_X$ is **continuous**

::: Question 

Can we define the Probability Mass Function for $X$?

:::


**Answer**: 

- Yes we can, but it would be useless - pmf carries no information
- This is because 
$$
f_X(x) = P(X=x) = 0 \,, \quad \forall \, x \in \mathbb{R}
$$




## Probability Mass Function? {.smaller}


- Indeed, for all $\varepsilon>0$ we have $\{ X = x \} \subset \{  x - \varepsilon < X \leq x  \}$
- Therefore by the properties of probabilities we have
\begin{align*}
P (X = x ) & \leq P(  x - \varepsilon < X \leq x ) \\
           & = P(X \leq x) - P(X \leq x - \varepsilon) \\
           & = F_X(x) - F_X(x-\varepsilon)
\end{align*}
where we also used the definition of $F_X$
- Since $F_X$ is continuous we get
$$
0 \leq P(X = x) \leq \lim_{\varepsilon \to 0} F_X(x) - F_X(x-\varepsilon) = 0
$$
- Then $f_X(x) = P(X=x) = 0$ for all $x \in \mathbb{R}$






## Probability Density Function {.smaller}

**pmf** carries no information for continuous RV -- We instead define the **pdf**


::: Definition

The **Probability Density Function** or **pdf** of a continuous rv $X$ is
a function $f_X \colon \mathbb{R} \to \mathbb{R}$ s.t.
$$
F_X(x) =  \int_{-\infty}^x f_X(t) \, dt  \,, \quad \forall \, x \in
\mathbb{R}
$$

:::


**Technical issue**: 

- If $X$ is continuous then pdf **does not exist** in general  
(absolute continuity is required)
- Counterexamples are rare, therefore we will **assume** existence of pdf




## Probability Density Function {.smaller}
### Properties

::: Proposition

Suppose $X$ is continuous rv. They hold

- The **cdf** $F_X$ is continuous and differentiable (a.e.) with
$$
F_X' = f_X
$$

- Probability can be computed via
    $$
    P(a \leq X \leq b) = \int_{a}^b f_X (t) \, dt \,, \quad \forall \, a,b \in 
    \mathbb{R} \,, \,\, a \leq b 
    $$

:::





## Example - Logistic Distribution {.smaller}


- The random variable $X$ has **logistic distribution** if its **pdf** is
$$
f_X(x) = \frac{e^{-x}}{(1+e^{-x})^2}
$$



```{r}

# Define the probability density function (PDF)
f_X <- function(x) {
  exp(-x) / (1 + exp(-x))^2
}

# Generate x values for plotting
x_values <- seq(-10, 10, length.out = 1000)

# Calculate corresponding y values
y_values <- f_X(x_values)

# Plot the distribution using base R plot function
plot(x_values, 
    y_values, 
    type = "l", 
    col = "blue", 
    lwd = 4,
    xlab = "", 
    ylab = "")

mtext("x", side=1, line=3, cex=2)
mtext("f_X", side=2, line=2.5, cex=2)


# Add a grid for better visualization
grid()

```



## Example - Logistic Distribution {.smaller}


- The random variable $X$ has **logistic distribution** if its **pdf** is
$$
f_X(x) = \frac{e^{-x}}{(1+e^{-x})^2}
$$



- The **cdf** can be computed to be
$$
F_X(x) = \int_{-\infty}^x f_X(t) \, dt = \frac{1}{1+e^{-x}}
$$


- The RHS is known as **logistic function**





## Example - Logistic Distribution {.smaller}


**Application**: Logistic function models expected score in chess (see [Wikipedia](https://en.wikipedia.org/wiki/Elo_rating_system))



- $R_A$ is ELO rating of player $A$, $R_B$ is ELO rating of player $B$
- $E_A$ is expected score of player $A$: 
$$
E_A := P(A \text{ wins}) + \frac12 P(A \text{ draws})
$$
- $E_A$ modelled by logistic function
$$
E_A := \frac{1}{1+ 10^{(R_B-R_A)/400} }
$$
- Example: Beginner is rated $1000$, International Master is rated $2400$
$$
R_{\rm Begin} = 1000, \quad  R_{\rm IM}=2400 , \quad E_{\rm Begin} = \frac{1}{1 + 10^{1400/400}} = 0.00031612779
$$




## Characterization of pmf and pdf {.smaller}


::: Theorem

Let $f \colon \mathbb{R} \to \mathbb{R}$. Then $f$ is pmf or pdf of a RV $X$ iff

1. $f(x) \geq 0$ for all $x \in \mathbb{R}$
2. $\sum_{x=-\infty}^\infty f(x) = 1 \,\,\,$ (pmf) $\quad$ or $\quad$ $\int_{-\infty}^\infty f(x) \, dx = 1\,\,\,$ (pdf)

:::

In the above setting: 

- The RV $X$ has **distribution**
$$
P(X = x) = f(x)  \,\,\, \text{ (pmf) } \quad  \text{ or }  \quad
P(a \leq X \leq b) = \int_a^b f(t) \, dt   \,\,\, \text{ (pdf)}
$$

- The symbol $X \sim f$ denotes that $X$ has distribution $f$





## Summary - Random Variables {.smaller}

- Given probability space $(\Omega, \mathcal{B}, P)$ and a Random Variable $X \colon \Omega \to \mathbb{R}$

- **Cumulative Density Function (cdf)**:
$F_X(x) := P(X \leq x)$


| Discrete RV                       |        Continuous RV         |
|--------------                     |----------------              |
| $F_X$ has **jumps**               |  $F_X$ is **continuous**     |
|**Probability Mass Function (pmf)**|**Probability Density Function (pdf)**|
| $f_X(x) := P(X=x)$                | $f_X(x) := F_X'(x)$          |  
| $f_X \geq 0$                      |  $f_X \geq 0$                |
|$\sum_{x=-\infty}^\infty f_X(x) = 1$| $\int_{-\infty}^\infty f_X(x) \, dx = 1$ |
| $F_X (x) = \sum_{k=-\infty}^x f_X(k)$| $F_X (x) = \int_{-\infty}^x f_X(t) \, dt$
| $P(a \leq X \leq b) = \sum_{k = a}^{b} f_X(k)$ | $P(a \leq X \leq b) = \int_a^b f_X(t) \, dt$ |




# Part 3: <br>Expected value {background-color="#cc0164" visibility="uncounted"}

::: footer

<div color="#cc0164">  </div>

:::





## Functions of Random Variables {.smaller}

- $X \colon \Omega \to \mathbb{R}$ random variable and $g \colon \mathbb{R} \to \mathbb{R}$ function
- Then $Y:=g(X) \colon \Omega \to \mathbb{R}$ is random variable
- For $A \subset \mathbb{R}$ we define the pre-image 
$$
g^{-1}(A) := \{  x \in \mathbb{R} \colon g(x) \in A \}
$$
- For $A=\{y\}$ single element set we denote
$$
g^{-1}(\{y\}) = g^{-1}(y) =  \{  x \in \mathbb{R} \colon g(x) = y \}
$$
- The distribution of $Y$ is
$$
P(Y \in A) = P(g(X) \in A ) = P(X \in g^{-1}(A))
$$



## Functions of Random Variables {.smaller}

**Question**: What is the relationship between $f_X$ and $f_Y$?

- **$X$ discrete**: Then $Y$ is discrete and
$$
f_Y (y) = P(Y = y) = \sum_{x \in g^{-1}(y)} P(X=x) = \sum_{x \in g^{-1}(y)} f_X(x) 
$$

- **$X$ and $Y$ continuous**: Then
\begin{align*}
F_Y(y) & = P(Y \leq y) = P(g(X) \leq y) \\
       & = P(\{ x \in \mathbb{R} \colon g(x) \leq y \}   ) = \int_{\{ x \in \mathbb{R} \colon g(x) \leq y \}} f_X(t) \, dt
\end{align*}





## Functions of Random Variables {.smaller}

**Issue**: The below set may be tricky to compute
$$
\{ x \in \mathbb{R} \colon g(x) \leq y \}
$$

However it can be easily computed if $g$ is **strictly monotone**:

- **g strictly increasing**: Meaning that 
$$
x_1 < x_2 \quad \implies \quad g(x_1) < g(x_2)
$$

- **g strictly decreasing**: Meaning that 
$$
x_1 < x_2 \quad \implies \quad g(x_1) > g(x_2)
$$

- In both cases $g$ is **invertible**





## Functions of Random Variables {.smaller}

Let $g$ be **strictly increasing**:

- Then
$$
\{ x \in \mathbb{R} \colon g(x) \leq y \} = \{ x \in \mathbb{R} \colon x \leq g^{-1}(y) \}
$$

- Therefore
\begin{align*}
F_Y(y) & = \int_{\{ x \in \mathbb{R} \colon g(x) \leq y \}} f_X(t) \, dt 
         = \int_{\{ x \in \mathbb{R} \colon x \leq g^{-1}(y) \}} f_X(t) \, dt  \\
       & = \int_{-\infty}^{g^{-1}(y)} f_X(t) \, dt 
         = F_X(g^{-1}(y)) 
\end{align*}





## Functions of Random Variables {.smaller}

Let $g$ be **strictly decreasing**:

- Then
$$
\{ x \in \mathbb{R} \colon g(x) \leq y \} = \{ x \in \mathbb{R} \colon x \geq g^{-1}(y) \}
$$

- Therefore
\begin{align*}
F_Y(y) & = \int_{\{ x \in \mathbb{R} \colon g(x) \leq y \}} f_X(t) \, dt 
         = \int_{\{ x \in \mathbb{R} \colon x \geq g^{-1}(y) \}} f_X(t) \, dt  \\
       & = \int_{g^{-1}(y)}^{\infty} f_X(t) \, dt 
         = 1 - \int_{-\infty}^{g^{-1}(y)}f_X(t) \, dt  \\
       & = 1 - F_X(g^{-1}(y)) 
\end{align*}





## Summary - Functions of Random Variables {.smaller}


- **$X$ discrete**: Then $Y$ is discrete and
$$
f_Y (y) = \sum_{x \in g^{-1}(y)} f_X(x) 
$$

- **$X$ and $Y$ continuous**: Then
$$
F_Y(y) = \int_{\{ x \in \mathbb{R} \colon g(x) \leq y \}} f_X(t) \, dt
$$

- **$X$ and $Y$ continuous** and
    * **$g$ strictly increasing**: $F_Y(y) = F_X(g^{-1}(y))$
    * **$g$ strictly decreasing**: $F_Y(y) = 1 - F_X(g^{-1}(y))$







## Expected Value {.smaller}

Expected value is the **average** value of a random variable


::: Definition

$X$ rv and $g \colon \mathbb{R} \to \mathbb{R}$ function. The **expected value** or **mean**
of $g(X)$ is $\Expect [g(X)]$

- If $X$ discrete
$$
\Expect [g(X)]:= \sum_{x \in \mathbb{R}} g(x) f_X(x) = \sum_{x \in \mathbb{R}} g(x) P(X = x)
$$

- If $X$ continuous
$$
\Expect [g(X)]:= \int_{-\infty}^{\infty} g(x) f_X(x) \, dx
$$

:::





## Expected Value  {.smaller}
### Properties


In particular we have^[These follow by taking $g(x)=x$ in previous definitions, so that $g(X) = X$]


- If $X$ discrete
$$
\Expect [X] = \sum_{x \in \mathbb{R}} x f_X(x) = \sum_{x \in \mathbb{R}} x P(X = x)
$$

- If $X$ continuous
$$
\Expect [X] = \int_{-\infty}^{\infty} x f_X(x) \, dx
$$








## Expected Value  {.smaller}
### Properties

::: Theorem

$X$ rv, $g,h \colon \R \to \R$ functions and $a,b,c \in \mathbb{R}$. The expected value is **linear**
\begin{equation} \tag{1}
\Expect [a g(X) + b h(X) + c] = a\Expect [g(X)] + b \Expect [h(X)] + c
\end{equation}
In particular
\begin{align} \tag{2}
\Expect[aX] & = a \Expect[X] \\
\Expect [c] & = c \tag{3}
\end{align}
:::





## Expected Value  {.smaller}
### Proof of Theorem

- Equation (2) follows from (1) by setting $g(x)=x$ and $b=c=0$

- Equation (3) follows from (1) by setting $a=b=0$

- To show (1), suppose $X$ is continuous and set $p(x):=ag(x)+bh(x)+c$
\begin{align*}
\Expect [ag(X) + & b h(X) + c]  = \Expect [p(X)] = \int_{\R} p(x) f_X(x) \, dx \\
& = \int_{\R} (ag(x) + bh(x) + c) f_X(x) \, dx  \\
& = a\int_{\R} g(x) f_X(x) \, dx  + b\int_{\R} h(x) f_X(x) \, dx +  c\int_{\R} f_X(x) \, dx  \\
& = a \Expect[g(X)] + b \Expect[h(X)] + c
\end{align*}

- If $X$ is discrete just replace integrals with series in the above argument







## Expected Value  {.smaller}
### Further Properties

Below are further properties of $\Expect$, which we do not prove

::: Theorem

Suppose $X$ and $Y$ are rv. The expected value is:

- **Monotone**: $X \leq Y  \quad \implies \quad \Expect [X] \leq  \Expect [Y]$

- **Non-degenerate**: $\Expect[|X|] = 0 \quad \implies \quad X = 0$

- $X=Y \quad \implies \quad \Expect[X]=\Expect[Y]$ 

:::






## Variance {.smaller}

Variance measures how much a rv $X$ deviates from $\Expect[X]$

::: Definition
### Variance

The **variance** of a random variable $X$ is
$$
\Var[X]:= \Expect[(X - \Expect[X])^2]
$$

:::


**Note**:

- $\Var [X] = 0 \quad \implies \quad (X - \Expect[X])^2 = 0 \quad  \implies \quad X = \Expect [X]$
- If $\Var[X]$ is small then $X$ is close to $\Expect[X]$
- If $\Var[X]$ is large then $X$ is very variable




## Variance {.smaller}
### Equivalent formula

::: Proposition
$$
\Var[X] =  \Expect[X^2] - \Expect[X]^2
$$
:::

**Proof:**
\begin{align*}
\Var[X] & = \Expect[(X - \Expect[X])^2] \\
        & = \Expect[X^2 - 2 X \Expect[X] +  \Expect[X]^2] \\  
        & = \Expect[X^2] - \Expect[2 X \Expect[X]]  + \Expect[ \Expect[X]^2] \\
        & = \Expect[X^2] - 2 \Expect[X]^2 + \Expect[X]^2 \\
        & = \Expect[X^2] - \Expect[X]^2
\end{align*}





## Variance {.smaller}
### Variance is quadratic

::: Proposition

$X$ rv and $a,b \in \R$. Then
$$
\Var[a X + b] = a^2 \Var[X] 
$$

:::


**Proof:** Using linearity of $\Expect$ and the fact that $\Expect[c]=c$ for constants:
\begin{align*}
\Var[a X + b] & = \Expect[ (aX + b)^2 ] - \Expect[ aX + b ]^2 \\
              & = \Expect[ a^2X^2 + b^2 + 2abX ] - ( a\Expect[X]  + b)^2 \\
              & = a^2 \Expect[ X^2 ] + b^2 + 2ab \Expect[X] - 
                  a^2 \Expect[X]^2   - b^2 - 2ab \Expect[X] \\
              & = a^2 ( \Expect[ X^2 ] - \Expect[ X ]^2 ) = a^2 \Var[X]
\end{align*}





## Variance {.smaller}
### How to compute the Variance

We have
$$
\Var[X] =  \Expect[X^2] - \Expect[X]^2
$$

- $X$ discrete:
$$
E[X] = \sum_{x \in \mathbb{R}} x f_X(x) \,, \qquad 
E[X^2] = \sum_{x \in \mathbb{R}} x^2 f_X(x)
$$


- $X$ continuous:
$$
E[X] = \int_{-\infty}^\infty x f_X(x) \, dx \,, \qquad 
E[X^2] = \int_{-\infty}^\infty x^2 f_X(x) \, dx
$$





## Example - Gamma distribution {.smaller}
### Definition


The **Gamma distribution** with parameters $\alpha,\beta>0$ is
$$
f(x) := \frac{x^{\alpha-1} e^{-\beta{x}} \beta^{\alpha}}{\Gamma(\alpha)} \,, \quad x > 0
$$
where $\Gamma$ is the **Gamma function**
$$
\Gamma(a) :=\int_0^{\infty} x^{a-1} e^{-x} \, dx
$$





## Example - Gamma distribution {.smaller}
### Definition

**Properties of $\Gamma$**:

- The Gamma function coincides with the factorial on natural numbers
$$
\Gamma(n)=(n-1)! \,, \quad \forall \, n \in \N
$$

- More in general
$$
\Gamma(a)=(a-1)\Gamma(a-1) \,, \quad \forall \, a > 0
$$


-  Definition of $\Gamma$ implies normalization of the Gamma distribution:
$$
\int_0^{\infty} f(x) \,dx = 
\int_0^{\infty} \frac{x^{\alpha-1} e^{-\beta{x}} \beta^{\alpha}}{\Gamma(\alpha)}  \, dx
=  1 
$$



## Example - Gamma distribution {.smaller}
### Definition

$X$ has Gamma distribution with parameters $\alpha,\beta$ if 

- the pdf of $X$ is
$$
f_X(x) = 
\begin{cases} 
\dfrac{x^{\alpha-1} e^{-\beta{x}} \beta^{\alpha}}{\Gamma(\alpha)}  & \text{ if } x > 0 \\
0                                                                 & \text{ if } x \leq 0
\end{cases}
$$

- In this case we write $X \sim \Gamma(\alpha,\beta)$

- $\alpha$ is **shape** parameter

- $\beta$ is **rate** parameter




## Example - Gamma distribution {.smaller}
### Plot

Plotting $\Gamma(\alpha,\beta)$ for parameters $(2,1)$ and $(3,2)$

```{r}
# Set the shape and rate parameters for the first gamma distribution
alpha1 <- 2
beta1 <- 1

# Set the shape and rate parameters for the second gamma distribution
alpha2 <- 3
beta2 <- 2

# Generate values for the x-axis
x_values <- seq(0, 15, by = 0.1)

# Calculate the probability density function (PDF) values for each x for the first distribution
pdf_values1 <- dgamma(x_values, shape = alpha1, rate = beta1)

# Calculate the probability density function (PDF) values for each x for the second distribution
pdf_values2 <- dgamma(x_values, shape = alpha2, rate = beta2)

# Plot the gamma PDFs
plot(x_values, 
    pdf_values1, 
    type = "l", 
    col = "blue", lwd = 2,
    xlab = "", 
    ylab = "",
    ylim = c(0, max(pdf_values1, pdf_values2) + 0.1))

lines(x_values, 
      pdf_values2, 
      col = "red", 
      lwd = 2)

mtext("x", side=1, line=3, cex=2)
mtext("pdf", side=2, line=2.5, cex=2)

# Add a legend
legend("topright", legend = c(paste("Gamma(", alpha1, ",", beta1, ")", sep = ""),
                              paste("Gamma(", alpha2, ",", beta2, ")", sep = "")),
       col = c("blue", "red"), lty = 1, lwd = 2, cex = 1.5)

```  





## Example - Gamma distribution {.smaller}
### Expected value

Let $X \sim \Gamma(\alpha,\beta)$. We have:
\begin{align*}
\Expect [X] & = \int_{-\infty}^\infty x f_X(x) \, dx  \\
            & = \int_0^\infty  x \, \frac{x^{\alpha-1} e^{-\beta{x}} \beta^{\alpha}}{\Gamma(\alpha)} \, dx \\
            & = \frac{ \beta^{\alpha} }{ \Gamma(\alpha) }
            \int_0^\infty  x^{\alpha} e^{-\beta{x}}  \, dx
\end{align*}



## Example - Gamma distribution {.smaller}
### Expected value

Recall previous calculation:
$$
\Expect [X] = \frac{ \beta^{\alpha} }{ \Gamma(\alpha) }
            \int_0^\infty  x^{\alpha} e^{-\beta{x}}  \, dx
$$
Change variable $y=\beta x$ and recall definition of $\Gamma$:
\begin{align*}
 \int_0^\infty  x^{\alpha} e^{-\beta{x}}  \, dx & = 
  \int_0^\infty  \frac{1}{\beta^{\alpha}} (\beta x)^{\alpha} e^{-\beta{x}} \frac{1}{\beta} \, \beta \, dx \\
  & = \frac{1}{\beta^{\alpha+1}} \int_0^\infty  y^{\alpha} e^{-y} \, dy \\
  & = \frac{1}{\beta^{\alpha+1}} \Gamma(\alpha+1)
\end{align*}




## Example - Gamma distribution {.smaller}
### Expected value

Therefore
\begin{align*}
\Expect [X] & = \frac{ \beta^{\alpha} }{ \Gamma(\alpha) }
            \int_0^\infty  x^{\alpha} e^{-\beta{x}}  \, dx \\
            & =  \frac{ \beta^{\alpha} }{ \Gamma(\alpha) } \, \frac{1}{\beta^{\alpha+1}} \Gamma(\alpha+1) \\
            & = \frac{\Gamma(\alpha+1)}{\beta \Gamma(\alpha)}
\end{align*}

Recalling that $\Gamma(\alpha+1)=\alpha \Gamma(\alpha)$:
$$
\Expect [X] = \frac{\Gamma(\alpha+1)}{\beta \Gamma(\alpha)} = \frac{\alpha}{\beta}
$$



## Example - Gamma distribution {.smaller}
### Variance

We want to compute
$$
\Var[X] = \Expect[X^2] - \Expect[X]^2
$$

- We already have $\Expect[X]$
- Need to compute $\Expect[X^2]$



## Example - Gamma distribution {.smaller}
### Variance

Proceeding similarly we have:

\begin{align*}
\Expect[X^2] & = \int_{-\infty}^{\infty} x^2 f_X(x) \, dx \\
             & = \int_{0}^{\infty}  x^2 \, \frac{ x^{\alpha-1} \beta^{\alpha} e^{- \beta x} }{ \Gamma(\alpha) } \, dx \\
             & =  \frac{\beta^{\alpha}}{\Gamma(\alpha)} \int_{0}^{\infty}  x^{\alpha+1} e^{- \beta x} \, dx 
\end{align*}




## Example - Gamma distribution {.smaller}
### Variance

Recall previous calculation:
$$
\Expect [X^2] =  \frac{\beta^{\alpha}}{\Gamma(\alpha)} \int_{0}^{\infty}  x^{\alpha+1} e^{- \beta x} \, dx 
$$
Change variable $y=\beta x$ and recall definition of $\Gamma$:
\begin{align*}
 \int_0^\infty  x^{\alpha+1} e^{-\beta{x}}  \, dx & = 
  \int_0^\infty  \frac{1}{\beta^{\alpha+1}} (\beta x)^{\alpha + 1} e^{-\beta{x}} \frac{1}{\beta} \, \beta \, dx \\
  & = \frac{1}{\beta^{\alpha+2}} \int_0^\infty  y^{\alpha + 1 } e^{-y} \, dy \\
  & = \frac{1}{\beta^{\alpha+2}} \Gamma(\alpha+2)
\end{align*}



## Example - Gamma distribution {.smaller}
### Variance

Therefore
$$
\Expect [X^2]  = \frac{ \beta^{\alpha} }{ \Gamma(\alpha) }
            \int_0^\infty  x^{\alpha+1} e^{-\beta{x}}  \, dx 
             =  \frac{ \beta^{\alpha} }{ \Gamma(\alpha) } \, \frac{1}{\beta^{\alpha+2}} \Gamma(\alpha+2) 
              = \frac{\Gamma(\alpha+2)}{\beta^2 \Gamma(\alpha)}
$$
Now use following formula twice $\Gamma(\alpha+1)=\alpha \Gamma(\alpha)$:
$$
\Gamma(\alpha+2)= (\alpha + 1) \Gamma(\alpha + 1) = (\alpha + 1) \alpha \Gamma(\alpha)
$$
Substituting we get
$$
\Expect [X^2] = \frac{\Gamma(\alpha+2)}{\beta^2 \Gamma(\alpha)} = \frac{(\alpha+1) \alpha}{\beta^2}
$$



## Example - Gamma distribution {.smaller}
### Variance

Therefore
$$
\Expect [X] = \frac{\alpha}{\beta}  \quad \qquad
\Expect [X^2] = \frac{(\alpha+1) \alpha}{\beta^2}
$$
and the variance is
\begin{align*}
\Var[X] & = \Expect [X^2] - \Expect [X]^2 \\
        & = \frac{(\alpha+1) \alpha}{\beta^2} - \frac{\alpha^2}{\beta^2} \\
        & = \frac{\alpha}{\beta^2}
\end{align*}






## References